# Spectral Analysis in R

Recall that in spectral analysis we step into a "bizarro world", switching from the time domain to the frequency domain. There, two essential features are of interest: peaks, and background. In this tutorial, we will perform spectral analysis on Rio Grande streamflow, and see if we can learn anything interesting about it in this way.

### Packages

```{r, message=FALSE}
library(tidyverse)
library(scales)
library(astrochron)
library(lomb)
library(biwavelet)
library(scales)
```

### Dataset    

We will be looking at the daily discharge of the Rio Grande, which has been measured at Embudo, NM since 1889. The data and their associated metadata may be retrieved from the [USGS website](https://waterdata.usgs.gov/nwis/dv?cb_00060=on&format=rdb&site_no=08279500&legacy=&referred_module=sw&period=&begin_date=1889-01-01&end_date=2024-05-20). Let's load them and have a look:

```{r}
df <- read.table('https://waterdata.usgs.gov/nwis/dv?cb_00060=on&format=rdb&site_no=08279500&legacy=&referred_module=sw&period=&begin_date=1889-01-01&end_date=2024-05-20', skip=35, sep="\t")
names(df) <- c('agency', 'site_no', 'datetime','discharge (cf/s)','code')
df$datetime <- lubridate::as_datetime(df$datetime)
head(df)
```

First, let us inspect the distribution of values:

```{r}
hist(df$`discharge (cf/s)`, main = "Distribution of Rio Grande Discharge", xlab = "Discharge (cf/s)")
```

Like many other variables, streamflow is **positively skewed**: the distribution is asymmetric, with most values near zero and few instances of very high streamflow. Now let's inspect the temporal behavior:

```{r}
ggplot(df, aes(x = datetime, y = `discharge (cf/s)`)) +
  geom_line() +
  labs(x = "Time", y = "Discharge (cf/s)")
```

## Aggregate to monthly

```{r}
discharge_monthly <- df |> 
  group_by(Date = floor_date(ymd(df$datetime), '1 month')) |> 
  summarise(discharge = mean(`discharge (cf/s)`, na.rm = TRUE), .groups = 'drop')


ggplot(discharge_monthly, aes(x=Date, y=discharge)) +
  labs(title = "Rio Grande at Embudo, NM (monthly)",
       x="Year (CE)",
       y="dicharge (cf/s)") +
  geom_line() +
  ggtitle("Rio Grand Discharge") +
  theme_light()

```

### Even sampling

```{r}
missing_vals <- discharge_monthly$Date[which(is.na(discharge_monthly$discharge))]
missing_vals

df3 <- discharge_monthly |> 
  dplyr::filter(Date > max(missing_vals))

hist(as.numeric(diff(df3$Date)),main = "Distribution of Time Steps", xlab = "Days")

df4 <- df3 |> 
  mutate(Date = decimal_date(Date)) |> 
  astrochron::linterp(dt=(1/12),genplot = F) |> 
  dplyr::filter(Date > max(missing_vals))

ggplot(df4, aes(x=Date, y=discharge)) +
  labs(title = "Rio Grande at Embudo, NM (30.4375-day period)",
       x="Year (CE)",
       y="dicharge (cf/s)") +
  geom_line() +
  theme_light()
```

### multi-taper

```{r}
mtm1 <- mtm(df4,output = 1,verbose = F) |> 
  mutate(Period = 1/Frequency,
                  Power = Power/1000) |> #account for differing units in astrochrons MTM
  dplyr::select(Period, Power)

reverselog_trans <- function(base = exp(1)) {
  trans <- function(x) -log(x, base)
  inv <- function(x) base^(-x)
  trans_new(paste0("reverselog-", format(base)), trans, inv, 
            log_breaks(base = base), 
            domain = c(1e-100, Inf))
}

ggplot(mtm1, aes(x=Period, y=Power)) +
  labs(title = "Rio Grande discharge spectrum (mtm)") +
  geom_line() +
  scale_y_log10() + 
  scale_x_continuous(trans = reverselog_trans(10),
                     breaks = c(100,50,20,10,5,2,1,0.5,0.2),
                     limits = c(100,0.2)) +
  theme_light()
```
The prominent feature is a strong annual cycle and higher-order harmonics, super-imposed on a “warm colored” background (i.e. variations at long timescales are larger than variations at short timescales). There are hints of long-term scaling as well, one in the subannual range, and one from period of 1 to 50y. We may be interested in the shape of this background, and whether it can be fit by one or more power laws.

In addition, we may be interested in interannual (year-to-year) to interdecadal (decade-to-decade) variations in streamflow. We would like to know whether the peak observed around 3-4 years in the plot above is *significant* with respect to a reasonable null.

## STL

To do both of these things it would make sense to remove the annual cycle. We'll use STL decomposition for this.

### Gaussianize

Now, because of the positive skew mentioned above, it turns out that the STL decomposition has trouble with this dataset. Instead, we can use the gaussianize() to map this dataset to a standard normal. There are applications for which this would be undesirable, but for this purpose it is good:

Gaussianize is available as part of the geoChronR package - but it's also pretty simple, so let's just define that here: 

```{r}
gaussianize <- function(X) {
    if (!is.matrix(X)) {
        X = as.matrix(X)
    }
    p = NCOL(X)
    n = NROW(X)
    Xn = matrix(data = NA, nrow = n, ncol = p)
    for (j in 1:p) {
        nz <- which(!is.na(X[, j]))
        N <- length(nz)
        R <- rank(X[nz, j])
        CDF <- R/N - 1/(2 * N)
        Xn[nz, j] <- qnorm(CDF)
    }
    return(Xn)
}

```

### MTM, again

Let's take a look at how this affects the spectrum:

```{r}
df5 <- df4 |> 
  mutate(discharge = gaussianize(discharge))
  
  
mtm2 <-   mtm(df5,output = 1,verbose = F) |> 
  mutate(Period = 1/Frequency, #Calculate frequency
         Power = Power/1000, #account for differing units in astrochrons MTM
        AR1_95_power = AR1_95_power/1000) |>  #account for differing units in astrochrons MTM
  dplyr::select(Period, Power)

ggplot(mtm2, aes(x=Period, y=Power)) +
  labs(title = "Rio Grande discharge spectrum (mtm)") +
  geom_line() +
  scale_y_log10() + 
  scale_x_continuous(trans = reverselog_trans(10),
                     breaks = c(100,50,20,10,5,2,1,0.5,0.2),
                     limits = c(100,0.2)) +
  theme_light()
```

### Applying STL

We see that the spectrum is nearly unchanged, but now we can apply STL:

```{r}
dis <- ts(as.numeric(df5$discharge), start = c(1889, 1), frequency = 12)
stl_res <- stl(dis, s.window = "periodic")
plot(stl_res)
```

### The trend

Now let's apply spectral analysis to the trend component:

```{r}
df6 <- df5 |> 
  mutate(discharge = as.numeric(stl_res$time.series[,2]))


mtm3 <-   mtm(df6,output = 1,verbose = F,)  |> 
  mutate(Period = 1/Frequency,
         Power = Power/1000, #account for differing units in astrochrons MTM
         AR1_95_power = AR1_95_power/1000) #account for differing units in astrochrons MTM
  

trendOnlySpectrum <- ggplot(mtm3, aes(x=Period, y=Power)) +
  labs(title = "Rio Grande discharge trend-only spectrum (mtm)") +
  geom_line() +
  geom_line(aes(y = AR1_95_power),color = "red") +
  scale_y_log10() + 
  scale_x_continuous(trans = reverselog_trans(10),
                     breaks = c(100,50,20,10,5,2,1,0.5,0.2),
                     limits = c(100,0.2)) +
  theme_light()

trendOnlySpectrum
```

Indeed this removed the annual cycle and its harmonics (for the most part). It's a fairly aggressive treatment, but we can now compare the significance of the interannual peaks w.r.t to an AR(1) benchmark. 

We see that the broadband peak in the 2-10y range is above the AR(1) spectrum, so these frequencies could be considered significant.

## Estimation of scaling behavior

In this last example, we fit a single scaling exponent to the timeseries, spanning the entire range of frequencies (periods). Under the hood, all we are doing is fitting a straight line to the spectrum:

```{r}
# Log-transform both axes
toScale <- dplyr::filter(mtm3, between(Period,0.2,100))
log_period <- log10(toScale$Period)
log_spec <- log10(toScale$Power)

# Fit a line
fit <- lm(log_spec ~ log_period)
toScale$plotLine <- predict(fit)
# Plot

trendOnlySpectrum + 
  geom_line(data = toScale,aes(x = Period,y = 10^plotLine),color = "blue",linetype = 3)


```

This results in a fairly steep exponent near `r round(coef(fit)[2], 2)`. If we were specifically interested in the scaling exponent (spectral slope) between periods of 2-100y, you would do it like so:

```{r}
# Log-transform both axes
toScale <- dplyr::filter(mtm3, between(Period,2,100))
log_period <- log10(toScale$Period)
log_spec <- log10(toScale$Power)

# Fit a line
fitLong <- lm(log_spec ~ log_period)
toScale$plotLine <- predict(fitLong)
# Plot

trendOnlySpectrum + 
  geom_line(data = toScale,aes(x = Period,y = 10^plotLine),color = "blue",linetype = 3)

```

We see that this results in a much flatter line, i.e. a smaller scaling exponent around `r round(coef(fitLong)[2], 2)`.

## Gap-tolerant spectral analysis

### Lomb-Scargle

We return to the original series and apply a technique to obtain the spectrum, keeping gaps in the series, known as the Lomb-Scargle periodogram:
```{r}
#Let's average (or bin) the data into monthly intervals
dfBinned <- df |> 
  mutate(year = year(datetime),
         month = month(datetime)) |> 
  group_by(year,month) |> 
  summarize(discharge = mean(`discharge (cf/s)`,na.rm = TRUE)) |> 
  mutate(yearDecimal = year + month/12 - 1/24) |> 
  dplyr::filter(is.finite(discharge))

# Compute Lomb-Scargle periodogram
lomb <- lomb::lsp(x = dfBinned$discharge, 
                  times = dfBinned$yearDecimal, 
                  ofac = 4, 
                  scaleType = "period",normalize = "press",
                  plot = FALSE)

ltp <- data.frame(Period = 1/lomb$scanned, Power = lomb$power)
# Plot
 ggplot(ltp, aes(x=Period, y=Power)) +
  labs(title = "Rio Grande discharge Lomb-Scargle spectrum") +
  geom_line() +
  #geom_line(aes(y = AR1_95_power),color = "red") +
  scale_y_log10() + 
  scale_x_continuous(trans = reverselog_trans(10),
                     breaks = c(100,50,20,10,5,2,1,0.5,0.2),
                     limits = c(100,0.2)) +
  theme_light()

```

We can see that this resembles our interpolated MTM approach - but has some differences. Let's plot on the same graph to take a closer look:

```{r}
comboPlotData <- bind_rows(mutate(mtm1,method = "mtm"),
                           mutate(ltp,method = "lomb-scargle"))
  
  ggplot(comboPlotData, aes(x=Period, y=Power,color = method)) +
  labs(title = "Rio Grande discharge spectrum") +
  geom_line() +
  scale_y_log10() + 
  scale_x_continuous(trans = reverselog_trans(10),
                     breaks = c(100,50,20,10,5,2,1,0.5,0.2),
                     limits = c(100,0.2)) +
  theme_light()

```
It is often useful to be able to compare methods and/or parameter choices, to see if the results are robust. We see that some choices lump peaks into broad bands, others tend to slice them up.

### Wavelet

Wavelet analysis using the Morlet wavelet:

```{r}
# Convert to matrix format required by biwavelet
dat <- cbind(time_vec, dis_vec)
dat <- dat[1:1880,]

# Compute wavelet transform
wav <- biwavelet::wt(dat)

# Plot wavelet power spectrum
biwavelet::plot.biwavelet(wav, plot.phase = FALSE, type = "power.norm")
```


We'll explore wavelets more in the next chapter.

## Takeaways

There are many methods for looking at the spectra of time series. For unevenly sampled series, we are more limited in methods, and interpolation may cause artifacts in our analyses.

